Do Drivers Dream of Walking? An Investigation of Travel Mode Dissonance from the Perspective of Subjective Wellbeing

### 2019-11-01 Dissonance_Santiago_v1.tex
Added discussion based on the files sent by Beatriz, and prepared files for submission

  \begin{abstract}
  Transportation in most of the world has been dominated for decades by a
  fascination with the automobile. Nonetheless, there is an increasing
  recognition that to achieve a variety of economic, environmental, and
  public health policy goals it is important to attract and retain active
  travelers and users of public transportation. A challenge with the way
  transportation policy is developed, however, is that it tends to focus
  on utilitarian considerations, which may miss other potential aspects of
  transportation that users value. In particular, subjective wellbeing
  (SWB) has been proposed as a way to enhance our understanding of the
  preferences and choices of travellers, as well as a way to evaluate the
  benefits of transportation beyond utilitarian considerations. The
  objective of this paper is to analyze the modes that people commonly
  use, and to what extent they are aligned (or not) with a variety of
  affective values associated with SWB. In other words, we are interested
  in the potential for dissonance with respect to the primary mode of
  travel, from the perspective of affective values. The study is based on
  data collected from a sample of travellers in the city of Santiago, in
  Chile. Participants in the study were asked about their usual mode of
  travel, and then were asked to name the mode or modes that they
  associate with the affective values of freedom, enjoyment, happiness,
  poverty, luxury and status. The results indicate that users of public
  transportation experience the most dissonance in terms of affective
  values, and active travellers the least. For those travellers who
  experience dissonance, active travel is the mode most commonly
  associated with freedom, enjoyment, and happiness, public transportation
  is most commonly associated with poverty, and the automobil is most
  commonly associated with luxury and status.
  \end{abstract}

@@

A useful lens to investigate the social and psychological functions of
transportation is by means of the concept of subjective wellbeing (SWB;
see, inter alia, De Vos et al., 2013; and Chatterjee et al., 2019). SWB
is defined by the OECD as ``{[}g{]}ood mental states, including all of
the various evaluations, positive and negative, that people make of
their lives, and the affective reactions of people to their
experiences'' (\textbf{OECD, 2013, p.~10}). As memorably put by Steg
(2005): is the use of car a must, or is it a lust? A close alignment, or
consonance, between affective values and the mode of transportation used
to travel can result in greater subjective wellbeing and increase the
probability of choosing that mode (i.e., by satisfying the \emph{lust});
on the other hand, dissonance, that is, the lack of concordance between
the mode used and affective values would be detrimental to subjective
wellbeing, especially of those for whom the use of the mode is a
\emph{must} (De Vos, 2019; Mokhtarian and Pendyala, 2018). With these
considerations in mind, one aim of the present research is to
investigate \emph{who} experiences dissonance with respect to their
primary mode of travel and affective values, namely freedom, enjoyment,
happiness, feelings of poverty, luxury, and status. Furthermore, it is
possible, even if use of a particular mode is a \emph{must}, that
travellers may still \emph{lust} for something else - in other words,
the grass may in fact look greener from the window of the car. For this
reason, we also aim to investigate which mode or modes are most commonly
associated with affective values, with a focus on those travellers who
experience dissonance in their primary mode.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The research reported here contributes to an emerging literature on
  the topic of transportation and affective values in the context of the
  Global South (Al-Ayyash and Abou-Zeid, 2019; Bejarano et al., 2017;
  Shao and Liang, 2019; Van et al., 2014; Zorrilla et al., 2019); to the
  best of our knowledge, the case of Chile has not yet been reported.
\item
  Although there is an extensive literature on the enjoyment of commute
  and other affective values (see for instance Paez and Whalen, 2010;
  Redmond and Mokhtarian, 2001; Whalen et al., 2013; Ye and Titheridge,
  2017), from a hedonic and even eudaimonic perspectives the analysis
  has yet to be applied more fully in terms of distributional issues --
  i.e.~which groups more commonly experience dissonance (see De Vos,
  2018).
\item
  The analysis shows the attitudes of people towards their primary mode
  and their perception towards `ideal modes' -- implying their
  preferences, even in situations when their ideal mode is not part of
  their actual choice set. More concretely, the results indicate that
  users of public transportation experience the most dissonance in terms
  of affective values, and active travellers the least. For those
  travellers who experience dissonance, active travel is the mode most
  commonly associated with freedom, enjoyment, and happiness, public
  transportation is most commonly associated with poverty, and the
  automobil is most commonly associated with luxury and status. We also
  find that there are some substantial variations in dissonance by age,
  education, income, and typical commute time. The attitudes of
  travellers towards transport modes are critical factors to be
  considered by policy-makers in case they want to promote and increase
  the use of public transport or active modes (Bornioli et al., 2019; De
  Vos et al., 2019; De Vos and Witlox, 2017; Garling et al., 2019;
  Redman et al., 2013).
\item ~
  \hypertarget{background}{%
  \section{Background}\label{background}}
\end{enumerate}

@@

\hypertarget{data-and-methods}{%
\section{3. Data and Methods}\label{data-and-methods}}

\hypertarget{context}{%
\section{3.1 Context}\label{context}}

Words to describe Santiago go here.

\hypertarget{sample}{%
\section{3.2 Sample}\label{sample}}

@@

\hypertarget{who-experiences-dissonance}{%
\section{4.1 Who experiences
dissonance?}\label{who-experiences-dissonance}}

@@

\caption{\label{tab:table-cross-tabulation-results-without-instrumental}\label{tab:cross-tabulation-results}Percentage of respondents who report mode dissonance with respect to various SWB and instrumental factors}
\centering
\begin{tabular}[t]{lcccccccccccc}
\toprule
Variable & Freedom & $\chi^2$ p-val & Enjoyment & $\chi^2$ p-val & Happiness & $\chi^2$ p-val & Poverty & $\chi^2$ p-val & Luxury & $\chi^2$ p-val & Status & $\chi^2$ p-val\\
\midrule
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Age}}\\
\hspace{1em}< 35 & 89.88 &  & 94.33 &  & 93.52 &  & 87.63 &  & 89.43 &  & 87.17 & \\

@@

\textbf{FOR DISCUSSION}

Al-Ayyash and Abou-Zeid (2019) Bornioli et al. (2019) De Vos et al.
(2019) Gatersleben and Uzzell (2007) Smith (2017) Ye and Titheridge
(2017)

@@

The preceding analysis suggests that there is significant mode
dissonance along various dimensions and for various affective values.
Once we have developed a profile of the travelers who experience mode
dissonance, we are interested in what modes they tend to associate with
various affective values. For this analysis, we employ a visualization
technique known as a mosaic plot that is used for exploring categorical
data (Friendly, 1994; Hofmann, 2000). A mosaic plot is a visual
representation of a multi-way contingency table, whereby the dimensions
of each tile in the mosaic are proportional to the values in the
underlying table.

We begin by plotting the cross-tabulation of primary mode of travel to
mode associated with each affective value (see Figure
\ref{fig:mosaic-plots-by-attribute}). The height of tile \(j\) in these
mosaics corresponds to the proportion of travellers who primarily travel
by mode \(j\), hence we can see that the most common primary mode of
travel in this sample is by public transportation, followed by car,
active travel, and other (also see Figure
\ref{fig:primary-mode-travel}). Notice that there are some small
discrepancies in the heights across affective values: these are caused
by a small number of non-responses. Next, the width of the tile \(k\) in
the mosaics is proportional to the frequency with which a mode was named
in relation to affective value \(i\). Recall that respondents could
nominate more than one mode in response to each affective value, and
some did. For this reason, we used sampling weights, so that the total
number of responses corresponds to the size of the sample.

In terms of Freedom, we see that Car users frequently identify this
affect with the use of Car; however, those who do not, very often select
Active Travel as the mode that better represents Freedom. A majority of
active travellers, on the other hand, identify Active Travel as the mode
associated with Freedom; few of them attach this value to Car, and even
fewer to public transportation and other modes. Travellers whose primary
mode of travel is public transportation seldom associate this mode with
Freedom, and in fact more frequently relate this affective response to
Active Travel followed by Car. Users of other modes seem to be split
almost equally in their attribution of Freedom, between Active Travel,
Car, and Other. In general, Active Travel is the mode most often
identified as embodying Freedom, with the exception of Car travellers.

@@

\begin{quote}
\begin{quote}
\begin{quote}
Efectivamente lujo y estatus parecen similares a primera vista (Figura
1). En general, estos dos valores afectivos son asociados mucho mas
comunmente con el automovil. Por otro lado, hay algunas diferencias
interesantes cuando se les ve desde la perspectiva de diferentes
viajeros.
\end{quote}
\end{quote}
\end{quote}

\begin{quote}
\begin{quote}
\begin{quote}
Por ejemplo, la tendencia a asociar estatus con el automovil tiende a
incrementarse con la edad, con la posible excepcion de los usuarios de
transporte publico, para quienes otras opciones (posiblemente taxi),
ganan en estatus para los viajeros mayores (\textgreater54) a costa de
otros modos, incluido el automovil. En contraste, el automovil pierde en
el valor de lujo con la edad, mientras que otras opciones pasan a ser
percibidas como lujo mas comunmente con la edad, incluso para los
usuarios de automovil (Figure 2).
\end{quote}
\end{quote}
\end{quote}

\begin{quote}
\begin{quote}
\begin{quote}
se vuelven mas comunmente simbolos de estatus que cuando vemos la manera
en que viajeros de diferentes edades proyectan lujo y estatus en varios
modos, los usuarios de automovil mayores (\textgreater54) muestran menos
disonancia en status que en lujo, mientras que los viajeros activos
presentan mas disonancia en estatus que en lujo, y de hecho proyectan
estatus mas comunmente en el automovil.
\end{quote}
\end{quote}
\end{quote}

\begin{quote}
\begin{quote}
\begin{quote}
La manera en que el lujo se asocia de manera relativamente mas frecuente
con otras opciones que el estatus tambien se observa cuando vemos los
diferentes niveles de educacion. La excepcion parece ser los usuarios de
transporte activo, especialmente aquellos con educacion superior y de
post-grado, que asocian tanto estatus como lujo con el transporte activo
mas que los usuarios de otros modos.
\end{quote}
\end{quote}
\end{quote}

@@

Words go here to discuss results by age.

@@

Words go here to discuss results by education.

@@

Words go here to discuss results by income.

@@

Words go here to discuss results by time.

@@

\textbf{\emph{FOR DISCUSSION}}

De Vos (2018)

@@

Accordingly, the aim of this paper is to analyse some of the factors
that can potentially affect the levels of SWB in the wide spectrum of
public, private and active transport modes in the context of Santiago de
Chile. Previous research suggests that public transport users experience
a negative gap in desired and actual travel time, whereas active
travellers don't seem to mind somewhat longer trips than what strictly
needed (e.g., Páez and Whalen, 2010). In the context of a Latin American
country, historically the poor travel experience of public transit users
and active travellers reflects the social inequalities of social groups
that have no alternative but to move in these types of modes while
living far from the work centres and main activities. The paper aims to
reflect on the possibilities of increasing the SWB while travelling, and
how this could increase the attractiveness of public transport and
active travel in the socio-demographic groups studied. This can help not
only to reduce the use of private transport but also decrease the
inequality gap between these users.

The paper explores how positive affective and instrumental/utilitarian
factors of SWB correspond to the use of the primary mode of
transportation. It uses an analysis of correspondence that considers the
actual primary mode versus the mode that individuals associated to both
utilitarian (designed to be useful or practical rather than attractive)
and affective (expressing a person's feelings) SWB factors. For example,
it will be analysed if a person who uses public transport as the primary
mode, tends to associate these modes with positive affective factors (as
freedom, enjoyment or happiness) or positive utilitarian factors (as
security, time-savings or efficiency). We hypothesize that there is a
more persistent non-correspondence or mismatch within the more
disadvantage groups or within the users of public transport.

@@

>>>>

  \begin{abstract}
  Subjective wellbeing is a topic that has attracted considerable
  attention in recent years due to the way it correlates with health. From
  a transportation perspective, there is a burgeoning literature on the
  way travel can impact subjective wellbeing, and how this, in turn, can
  influence behavior. After decades of fascination with the automobile,
  subjective wellbeing is increasingly recognized as an alternative
  approach to assess transportation, and to gain valuable insights to
  attract and retain active travelers and users of public transportation.
  The objective of this paper is to analyze a number of affective values
  associated with subjective wellbeing and the modes of transportation
  that people commonly use. In particular, we are interested in the
  potential for dissonance with respect to the primary mode of travel. The
  study is based on data collected from a sample of travellers in the city
  of Santiago, in Chile. Participants in the study were asked about their
  usual mode of travel, and then were asked to name the mode or modes that
  they associate with the affective values of freedom, enjoyment,
  happiness, poverty, luxury and status. The results indicate that users
  of public transportation experience the most dissonance in terms of
  affective values, and active travellers the least. For those travellers
  who experience dissonance, active travel is the mode most commonly
  associated with freedom, enjoyment, and happiness, public transportation
  is most commonly associated with poverty, and the automobil is most
  commonly associated with luxury and status.
  \end{abstract}

@@

A useful lens to investigate the social and psychological functions of
transportation is by means of the concept of subjective wellbeing (SWB;
see, inter alia, De Vos et al., 2013; and Chatterjee et al., 2019). SWB
is defined by the OECD as ``{[}g{]}ood mental states, including all of
the various evaluations, positive and negative, that people make of
their lives, and the affective reactions of people to their
experiences'' (OECD, 2013, p. 10). As memorably put by Steg (2005): is
the use of car a must, or is it a lust? A close alignment, or
consonance, between affective values and the mode of transportation used
to travel can result in greater subjective wellbeing and increase the
probability of choosing that mode (i.e., by satisfying the \emph{lust});
on the other hand, dissonance, that is, the lack of concordance between
the mode used and affective values would be detrimental to subjective
wellbeing, especially of those for whom the use of the mode is a
\emph{must} (De Vos, 2019; Mokhtarian and Pendyala, 2018). With these
considerations in mind, one aim of the present research is to
investigate \emph{who} experiences dissonance with respect to their
primary mode of travel and affective values, namely freedom, enjoyment,
happiness, feelings of poverty, luxury, and status. Furthermore, it is
possible, even if use of a particular mode is a \emph{must}, that
travellers may still \emph{lust} for something else - in other words,
the grass may in fact look greener from the window of the car. For this
reason, we also aim to investigate which mode or modes are most commonly
associated with affective values, with a focus on those travellers who
experience dissonance in their primary mode.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  The research reported here contributes to an emerging literature on
  the topic of transportation and affective values in the context of the
  Global South (Al-Ayyash and Abou-Zeid, 2019; Bejarano et al., 2017;
  Shao and Liang, 2019; Van et al., 2014; Zorrilla et al., 2019); to the
  best of our knowledge, the case of Chile has not yet been reported.
\item
  Although there is an extensive literature on the enjoyment of commute
  and other affective values (see for instance Paez and Whalen, 2010;
  Redmond and Mokhtarian, 2001; Whalen et al., 2013; Ye and Titheridge,
  2017), from a hedonic and even eudaimonic perspectives the analysis
  has yet to be applied more fully in terms of distributional issues --
  i.e.~which groups more commonly experience dissonance (see De Vos,
  2018).
\item
  The analysis shows the attitudes of people towards their primary mode
  and their perception towards `ideal modes' -- implying their
  preferences, even in situations when their ideal mode is not part of
  their actual choice set. More concretely, the results indicate that
  users of public transportation experience the most dissonance in terms
  of affective values, and active travellers the least. For those
  travellers who experience dissonance, active travel is the mode most
  commonly associated with freedom, enjoyment, and happiness, public
  transportation is most commonly associated with poverty, and the
  automobil is most commonly associated with luxury and status. We also
  find that there are some substantial variations in dissonance by age,
  education, income, and typical commute time. The attitudes of
  travellers towards transport modes are critical factors to be
  considered by policy-makers in case they want to promote and increase
  the use of public transport or active modes (Bornioli et al., 2019; De
  Vos et al., 2019; De Vos and Witlox, 2017; Garling et al., 2019;
  Redman et al., 2013).
\end{enumerate}

The structure of the paper is as follows. After these introductory
remarks, we follow up with a background section that reviews the
literature. Next, we discuss the case study and data used in the
research. Then, the analysis and results are presented, before
concluding with some discussion and directions for future research.

\hypertarget{background}{%
\section{2. Background}\label{background}}

@@

\hypertarget{case-study-and-data}{%
\section{3. Case Study and Data}\label{case-study-and-data}}

\hypertarget{context}{%
\subsection{3.1 Context}\label{context}}

The case study to contextualize the discussion is Santiago de Chile, the
capital of one of the countries with the highest levels of inequality in
terms of socio-economic distribution in the world. These inequities have
manifested in a disproportionate expense in transport by the different
socioe-conomic sectors, a disproportion in travel times and distances
travelled. On the other hand, car ownership and use of public transport
in the capital are directly conditioned by strong spatial segregation --
the higher the income, the higher is the use of the automobile;
conversely, and the lower the income, the higher is the predisposition
to use public transport. Although transport is a reflection of deep
inequities in various dimensions, it has triggered an exceptional
discomfort and disagreement with the daily travel experience and quality
of life.

\hypertarget{sample}{%
\subsection{3.2 Sample}\label{sample}}

@@

\hypertarget{who-experiences-dissonance}{%
\subsection{4.1 Who experiences
dissonance?}\label{who-experiences-dissonance}}

@@

\caption{\label{tab:table-cross-tabulation-results-without-instrumental}\label{tab:cross-tabulation-results}Percentage of respondents who report mode dissonance with respect to various SWB factors}
\centering
\begin{tabular}[t]{lcccccccccccc}
\toprule
Variable & Freedom & $\chi^2$ p-val & Enjoyment & $\chi^2$ p-val & Happiness & $\chi^2$ p-val & Poverty & $\chi^2$ p-val & Luxury & $\chi^2$ p-val & Status & $\chi^2$ p-val\\
\midrule
\addlinespace[0.3em]
\multicolumn{13}{l}{\textbf{Age}}\\
\hspace{1em}< 35 & 89.88 &  & 94.33 &  & 93.52 &  & 87.63 &  & 89.43 &  & 87.17 & \\

@@

@@

The preceding analysis suggests that there is significant mode
dissonance along various dimensions and for various affective values.
This is for the most part in line with previous research, although by
examining different affective values individually instead of a summary
measure of wellbeing, we are able to differentiate better in terms of
how travellers respond to different affects. Less is known about the
values that travellers associate with modes \emph{other} than the one
they use. For this reason, after developing a profile of the travelers
who experience mode dissonance in the preceding section, we are now
interested in the responses of travellers with respect to the modes they
tend to associate with various affective values. For this analysis we
employ a visualization technique known as a mosaic plot. Mosaic plots
are used in the exploration of multivariate categorical data (Friendly,
1994; Hofmann, 2000), and they provide visual representations of
multi-way contingency tables. The elements of the mosaic plot are called
\emph{tiles}, and the dimensions of each tile are proportional to values
of variables in the underlying table.

We begin our analysis by plotting primary mode of travel and the modes
associated with each affective value (see Figure
\ref{fig:mosaic-plots-by-attribute}). The height of tile \(j\) in these
mosaics corresponds to the proportion of travellers who primarily travel
by mode \(j\), hence we can see that the most common primary mode of
travel in this sample is by public transportation, followed by car,
active travel, and other (also see Figure
\ref{fig:primary-mode-travel}). Notice that there are some small
discrepancies in the heights across affective values: these are caused
by a small number of non-responses. Next, the width of the tile \(k\) in
the mosaics is proportional to the frequency with which a mode was named
in relation to affective value \(i\). Recall that respondents could
nominate more than one mode in response to each affective value, and
some did. For this reason, we used sampling weights, so that the total
number of responses corresponds to the size of the sample.

In terms of Freedom, we see that Car users frequently identify this
affect with the use of Car; however, those who do not, very often select
Active Travel as the mode that better represents Freedom. A majority of
active travellers, on the other hand, identify Active Travel as the mode
that evokes feelings of Freedom; few of them attach this value to Car,
and even fewer to public transportation and other modes. Travellers
whose primary mode of travel is public transportation seldom associate
this mode with Freedom, and in fact more frequently relate this
affective response to Active Travel followed by Car. Users of other
modes seem to be split almost equally in their attribution of Freedom,
between Active Travel, Car, and Other. In general, Active Travel is the
mode most often identified as embodying Freedom, with the exception of
Car travellers.

@@

@@

There are some interesting differences when viewed from the perspective
of travellers of different ages (see Figure
\ref{fig:mosaic-plots-by-age}). For example, as seen above, active
travel is commonly associated with freedom, happiness, and enjoyment,
even by car users, but especially by users of public transportation and
other modes. However, when we break this down by age, we notice that
this tendency weakens as people age, and older travellers increasingly
associate these affective values to the car. Furthermore, the tendency
to associate status with the car tends to increase with age, with the
possible exception of public transport users, for whom other options
(possibly taxi), gain status for older travellers (\textgreater{} 54) at
the expense of other modes, including the car. In contrast, the car
loses in the value of luxury with age, while other options become more
commonly associated with feelings of luxury for older respondents, even
for car users (Figure 2). Older car users (\textgreater{} 54) show less
dissonance in status than in luxury, while active travellers present
more dissonance in status than in luxury - and in fact, they more
commonly project status in the car.

@@

Only two affective values were significant in the preceding analysis
when seen from the perspective of education, namely luxury and status.
As seen in Figure \ref{fig:mosaic-plots-by-education}, the perceptions
of luxury and status are fairly consistent when differentiated by
education. Most respondents tend to associate luxury and status to the
use of car, albeit this association is stronger for status than for
luxury. The second mode most often mentioned in relation to luxury is
``other'' likely due to the flexibility and convenience (but cost) of
using taxis. The exception in this case is the traveller with graduate
education who uses active modes. This type of respondent, while still
more likely to associate luxury and status with the car, more commonly
links active travel to these affects than travellers with lower levels
of education. Most public transport users tend to associate luxury and
status to the car, however, they associate active modes with status more
than the same group with respect to luxury.

@@

Three affective values were significant in the analysis by income;
perhaps not surprisingly, all three have an economic dimension. Figure
\ref{fig:mosaic-plots-by-income} reflects some important social
inequalities associated with the way users feel about transportation in
Santiago. When the affective perception of poverty is analyzed, the
image shows that generally public transport is the mode most commonly
associated with poverty. There are some interesting differences,
however, to be observed by income level. Whereas middle income users of
public transportation associate this mode to poverty most frequently,
this is not the case of high income users of public transportation, who
tend to link active travel to poverty with greater frequency. Similarly,
active travel is more often associated with poverty by low income
travellers than middle or high income travellers. A possible explanation
for this is that walking or cycling are related to the idea of a lack of
financial means to access other modes of transport. For high income car
users, on the other hand, public transportation may have connotations of
poverty due to its massive use, whereas active travel may be seen more
as a lifestyle alternative.

With respect to luxury and status, we see again that these two values
coincide more strongly in their association with the car. Moreover, we
also notice that these perceptions become increasingly frequent at
higher levels of income. Unlike users of other modes, the high-income
public transport user group tends to concentrate a broader association
of luxury and status factors with active transport, reinforcing the
notion that at higher levels of income, people may see active modes as a
lifestyle alternative. High income active travellers, although still
more likely to associate luxury and status to the car, are more likely
than users of other modes to associate these affects to active travel,
perhaps due to an awareness of the benefits of walking and using the
bicycle. Furthermore, high-income users of cars are more likely to
experience consonance with respect to status and luxury, in what might
be a form of self-congratulatory confirmation of success.

@@

The last dimension that we examine is typical commute time. Four
affective values were significant along this dimension in Table
\ref{tab:cross-tabulation-results}. Figure
\ref{fig:mosaic-plots-by-travel-time} shows the way travellers associate
different affective values to modes of transportation by length of
typical commute. The first couple of factors, freedom and enjoyment, are
associated with eudaimonic and hedonic factors of wellbeing, as
discussed earlier in the paper. On the other hand, the luxury and status
are defined mainly by a socio-economic component. As seen in the figure,
freedom and enjoyment are most frequently mentioned in relation to
active travel, followed by car. However, active travel is more
frequently associated with freedom and enjoyment by people who typically
undertake relatively short trips (\textless20 min) or travellers whose
typical trips are longer (\textgreater60 min). In this respect, active
travel may be seen as an ideal mode for trips, and as an enviable mode
by people whose trips are longer.

In the case of luxury and status factors, both follow a similar pattern,
despite variations in travel times. Despite the travel time exceeding 60
minutes, car users consider that their transport mode is mostly
associated with luxury and status. The users of active transport in the
sample tend to travel less than 20 minutes, and this is the same segment
that considers that the status and luxury factors correspond mainly to
the car and then to active transport. Looking at the four factors set
out in Figure \ref{fig:mosaic-plots-by-travel-time}, it could be
understood that the modes that are associated with intrinsically
positive characteristics such as freedom and enjoyment tend to use
public transport. On the contrary, factors such as luxury and status
tend to have a negative connotation, where the association with the car
and other modes of transport (such as the individual taxi and the
collective taxi) tends to dominate.

@@

@@

Subjective wellbeing is a topic that has attracted considerable
attention in recent years due to its relationship with health. In the
field of transportation, there is a large and growing body of literature
that documents the way transportation can impact subjective wellbeing.
Likewise, there is strong evidence that subjective wellbeing can
influence travel behavior. As the world tries to move from a culture
dominated by a century-long love affair with the automobile, there is a
pressing need to understand how travellers perceive different modes of
transportation from the lens of subjective wellbeing. Insights in this
regard could prove valuable to develop and implement plans and policies
to attract and retain users to healthier, more environmentally friendly
transportation options, in particular active travel and public
transportation. For this reason, understanding mode dissonance, the
extent to which the context of travellers differs from their
aspirations, is a worthwhile topic for research.

The premise of the research presented in this paper is that subjective
wellbeing is a composite construct of various affective values. In this
paper we investigated mode dissonance from the perspective of six
affective values associated with subjective wellbeing, namely freedom,
enjoyment, happiness, poverty, luxury, and status. The research
presented here was based on a sample of travellers in Santiago, the
capital of Chile. Participants in this research were asked about their
typical mode of travel, and then about the mode or modes that they
associate with each of the six affective values. Analysis using
hypothesis testing (tests of independence) and vizualization techniques
(mosaic plots) uncovered interesting patterns. Some of our findings are
well aligned with previous research; for example, active travellers
experience less dissonance than car users, and users of public
transportation experience the most dissonance of all. However, by
considering affective values separately instead of aggregating them into
a single indicator of subjective wellbeing, we manage to preserve some
granularity with respect to various responses. This turns out to be
important, because hedonic/eudaimonic values are more frequently related
to active travel (freedom, enjoyment, happiness), whereas poverty is
more frequently related to public transportation and active travel.
Luxury and status, on the other hand, are more frequently associated to
luxury and status.

Further delving into the question of which modes are associated with
these affective values, we find that there are important differences in
terms of the typical mode of travel. Active travellers experience
dissonance with relatively little frequently with respect to freedom,
enjoyment, and happiness, but when they do, they tend to attach positive
values to the car. Car users experience dissonance with respect to these
affects more frequently than active travellers, and when they do, they
strongly relate possitive hedonic/eudaimonic values to active travel. In
other words, it is possible that drivers dream of walking when it comes
to feelings especially of enjoyment and happiness. The other side of the
coin is also interesting. When it comes to affective values with a
stronger socio-economic flavor, such as poverty, luxury, and status, car
users tend to experience dissonance less frequently than users of other
modes. Active travellers, altough more resistant to the lure of the car
compared to users of public transportation, also tend to attach values
of luxury and status to the car when they experience dissonance.

An examination of these effects by age, level of education, level of
income, and typical trip duration reveals that some of the responses
become more pronounced. For instance, older people are less likely than
younger people to associate active travel with positive
hedonic/eudaimonic affects, and are more likely to attach these values
to the car. People with higher incomes are more atuned to the luxury and
status values of cars, whereas lower income people are more likely to
relate active travel to luxury and status.

Our results not only help to flesh out some ways in which mode
dissonance could play out from the perspective of different affects, but
also does so in the context of a Latin American country, a region where
historically there are deep-seated tabooes with respect to different
modes of transportation: the poor travel by public transportation and/or
are forced active travellers; where the rich enjoy the luxury of private
vehicles and/or are active travellers by choice. In this way, the paper
helps us to reflect on the ways the public experiences
transportation-related subjective wellbeing, and how this in turn could
be used to increase the attractiveness of public transport and active
travel in the socio-demographic groups studied.

With respect to opportunities for future research, a possible avenue for
extending the findings reported in this paper relates to the use of the
modes other than the typical mode of transportation. In a recent paper
that investigaed commute satisfaction for car users, Al-Ayyash and
Abou-Zeid (2019) considered three models: for current trip satisfaction,
remembered satisfaction while using public transport, and current
satisfaction using public transport. The findings suggest that low
service quality in public transportation can result in a generalised
negative perception, and that this perception is more difficult to
smooth if commuters do not regularly use public transport. Another
avenue for future research could be to consider the mix of modes
typically used. While in this paper the analysis focused on the primary
mode of transportation, many travellers experience more than one mode of
transportation in their daily activities. For this reason, considering
the multimodal component of travel would be interesting; for example,
future research could consider people who eventually arrive by bicycle
to the metro station or people that, after using a colectivo for part of
a trip, end their journey by bus.

@@

\leavevmode\hypertarget{ref-oecd2013oecd}{}%
OECD, 2013. OECD guidelines on measuring subjective well-being. OECD
Publishing.
